Visual narratives of the COVID-19 pandemic

### 2022-09-06 Review/Response.qmd

# AE

::: callout
## Overall Evaluation
Overall largely
This paper provides an interesting overview of graphics used to portray data on the COVID 19 pandemic, largely from popular media. It attempts to use ideas and principles of graphics design and human factors
studies of graphical perception to evaluate the effectiveness of these displays for the public and policy
makers.

All in all, I think it largely succeeds in these aims and would be a suitable and interesting article for JDSSV readers. However, the two reviewers make substantial suggestions and comments which need to be considered before it is ready for publication. I recommend acceptance, following suitable revision (together with a reply from the authors to these reviews).
:::

**Authors' response:** Specific reviewer comments are addressed below, here, we address additional discussions brought up by the AE in response to the reviewers' comments.

@@

::: callout
## Reviewer 2 - Principles

To the reviewer’s comment, I add that there needs to some clear statement regarding the match
between the design of a graphic and the capabilities of the intended audience in determining whether a
graphic display is successful for its intended purpose. This is more important than simple design rules.
:::

**Authors' response:** We have added a short paragraph addressing this in the data communication subsection of the introduction, and used the opportunity to state that we have primarily focused on charts intended for mass consumption rather than experts in public health or epidemiology.

::: callout
### Other comments: Proportional symbol maps

@@

I was very surprised to see no mention or illustration of anamorphic maps or cartograms, where the area of each geographical unit is transformed to represent the count (cases/deaths). I strongly suggest including something on this.
:::

**Authors' response:** We have added a cartogram to the discussion of proportional symbol maps and choropleths. Thank you for catching an obvious omission on our part.

@@

::: callout
### Other comments: Color scheme obstructions

Color scheme is mentioned only in regard to Fig 6, but there are other examples where the color scheme used could be noted, either positively or negatively.

:::

**Authors' response:** We have added additional analysis of color where we feel it is useful; Figure 12 is one case where we chose not to mention color in part because one of the authors (Susan) can't tell if it's effective or not - she is colorblind, and while we have trouble with streamgraphs in general, that is particularly true if color is actually meaningful.

@@

::: callout
### Other comments: Uncertainty

Uncertainty: In the list of aspects mentioned in the middle of P 2, I am struck by the absence of mention
of uncertainty in visual representations of COVID. Clearly, the data are complex as they are, but
understanding their uncertainty is also a topic that could be mentioned or discussed.

:::

**Authors' response:** We wholeheartedly agree - uncertainty should be mentioned in the introduction, though I can't think of any specific examples of addressing the uncertainty through visualization (as opposed to footnotes about data quality) in mainstream graphics which were circulated at the time. Still, that is a rather glaring omission from both our discussion and the collection of graphs we've accumulated as interesting.

@@

::: callout
### Other comments: Temporal narrative

In Section 3, The temporal narrative, there could usefully be some discussion, and perhaps an illustration
of the use of dynamic charts like the moving bubble chart, to liberate the X axis from time, using
animation or an interactive slider.

:::

**Authors' response:** We've added a moving bubble chart - again, this is an obvious omission. Thank you for bringing it to our attention. Interestingly, it was somewhat difficult to find a chart like this in a mainstream media outlet when searching through retrospectively. Part of that may be the fact that the web isn't great at keeping archival content, and many of these dashboards have been overwritten through successive modification. I suspect another part, however, is that moving bubble charts require a lot of effort to read and understand, and possibly were more common on social media and in the data vis community writ large than in media intended for the general public.

@@

::: callout
### Other comments: standardization

More prominence could be given to the thoughtful standardization that went into some of these
displays to provide comparability of representation across global and local geographies and over
multiple waves of the pandemic. For example, in the early days cases and deaths were tied to calendar
dates giving time series charts with different origins. The simple device (I believe from Burn-Murdoch of
the FT) of making these all start when N cases had been observed was a game changer. There are
probably other examples.

:::

**Authors' response:** There may be other examples, but this is a good one. We've added a comment to that effect in the discussion of time series data, but we also mention standardization w.r.t. area and population in the choropleth/cartogram section. While the topic probably deserves a more thorough treatment, it's not easy to slide it in and maintain the current structure of the paper, so we hope that this suffices.

# Reviewer 1

@@

::: callout
### How was the sample of publications selected? Under what criteria?

As it stands, the selection seems arbitrary and very limited. Why the Financial Times and The New York Times and not others? Why not analyzing visualizations from non-Western countries? If the authors want to talk about a “global narrative”, the selection of graphics should also be global. A comparative cross-cultural study would be pretty productive, as it might reveal similarities and differences of graphics from different regions.
:::

**Authors' response:** The authors are from the United States and Germany; while we participated in many discussions on global social media platforms such as Twitter, we do not have the global perspective (or even facility with sufficient languages) to undertake a review of a global sample of media relating to COVID. We have added a section which explicitly states how charts were selected and what our individual biases are with respect to geography and culture.

@@

I’d eliminate the reference to this non-principle and I’d stick to the other principle, which is truly advisable and normative; from the paper itself: “ensuring clear understanding by organising the graphics in such a way that the story of the data is told most effectively.”

This would need to be extended, though, and more clearly explained. Making a visualization understandable doesn’t depend just on organization. It also involves making choices about encodings, types of charts and maps and, more importantly (something that I think this paper overlooks) actually testing that the graphic actually conveys the intended information to the audience it’s supposed to inform. Anyone who works in visualization will tell you that simply applying “good design practices” to a visualization will never guarantee for certain that it is understood.

:::

**Authors' response:** We have modified the principles section to remove the reference to Tufte and the data/ink ratio. We have slightly reorganized the introduction section to add subsections, and have added an additional paragraph which discusses the series of choices made in the process of designing a data visualization. While the general guidelines for making these choices are somewhat outside the scope of this paper, as this is an article, and not a book, and proper treatment of these decisions in a broader context would require at least a book, we hope that highlighting the more general context surrounding the creation of data visualizations and explicitly defining the scope for this article provides some additional clarity.

@@

I believe that the commentary on the graphics the authors chose would need to be more extensive and detailed. Just to give an example, they compare choropleth maps to proportional symbol maps, but they don’t reflect on the many shortcomings of the latter, such as the usual overlap between the scaled symbols. This might be related to space limits in the journal, but I think that this is a paper that deserves to be at least double its length to be truly comprehensive and detailed enough.

The authors also make too many assumptions about how a graphic will be interpreted. This is a direct quote: “The advantage of these displays is that they are very simple and allow for viewers to gain an intuitive understanding of the data”. Where is the evidence for such an assertion? The fact that the authors —likely professionals who are very familiar with statistics and visualization— find a graphic intuitive doesn’t make that graphic intuitive to other audiences. (The authors themselves actually address this in the following section, about log scales, arguably the best one in the paper.)

:::

**Authors' response:** While the editor has cautioned us not to double the length of the paper, where possible we have attempted to address some obvious shortcomings in our initial discussion without unduly increasing the paper's length. This paper is in part a result of discussing COVID graphics in several groups and seminars with data visualization professionals and also with laypeople; as a result, while we cannot directly cite a source saying that a specific graphic is simple and intuitive, we can directly assess the visual complexity of the chart and we can distill down the series of discussions we've had about these topics over the course of the pandemic to arrive at what we hope are fairly useful conclusions. In an attempt to be a bit more precise, however, where we could find relevant discussions on social media, we have included citations of those discussions to provide some support for claims that we have made summarizing the views of others.

# Reviewer 2

@@

As is, it reads badly and the organization is such that its points get lost. Keep the illustrations to the points you wish to make. Distinguish betwen map types and their uses. Use A, B-heads as you reorder your argument.

And in the discussion ... what have we learned? Well, there are static and dynamic maps, the former most common in news presentations, the latter in web-based. A trivial but useful point. Sliders are a good example how the latter expands what is in the traditional news map.

:::

**Authors' response:** The organization of this paper is certainly not the same as that which is proposed, in part because we have taken a narrative view that explores the different communication narratives of visualizations during the pandemic. Because we focus on the narratives and look at the chart types employed in service of those narratives, organization necessarily deviates from a straightforward analysis of each chart type in sequence, but this is by design.

### 2022-09-18 JDSSV_manuscript_submitted.Rnw
Revision 1
Updates - final round of revisions?

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Visual narratives of the Covid-19 pandemic}
\Plaintitle{Visual narratives of the Covid-19 pandemic}
\Shorttitle{Visual narratives of Covid-19}

%% - \Abstract{} almost as usual
\Abstract{
  Covid-19 has sparked a worldwide interest in understanding the dynamic evolution of a pandemic and tracking the effectiveness of preventive measures and rules. For this reason, numerous media and research groups have produced comprehensive data visualisations to illustrate the relevant trends and figures. In this paper, we will look at a selection of Covid 19 data visualisations to evaluate and discuss the currently established visualisation tools in terms of their ability to provide a communication channel both within the data science team and between data analysts, domain experts and a general interested audience. Although there is no set catalogue of evaluation criteria for data visualisations, we will try to give an overview of the different core aspects of visualisation evaluation and their competing principles.
  }

@@

Over the past two years, several waves of Covid-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience. According to the WHO, almost 600 million confirmed cases and over 6.45 million deaths have been reported as of late August 2022 \footnote{\url{https://www.who.int/docs/default-source/coronaviruse/who_mou_30_august-2022.pdf}}. The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation. From the beginning of the pandemic, data on the number of infections and covid-related deaths has been published daily and made available to the public. Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus. %So there are several goals to be achieved by visualising Covid-19 data.
These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation the general principles remain the same: ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.

@@

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the Covid-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards\citep{kochWelcomeRevolutionCOVID192021}.

@@

At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. 
These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). 
Efforts to improve the communication of quantitative and statistical information date from the beginning mass media and the ability to create charts and tables from data systematically, and while there continue to be efforts made in this area, historical papers outline many of the same issues that we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
In other words, while our **technology** for recording, visually summarizing, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when **designing** charts which communicate data accurately and effectively.

@@

\subsection{Effective Data Communication}
A best practice for effective communication (graphical or otherwise) is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%, or using absolute rather than relative numbers. 
Presenting information in a visually appealing graphic form and highlighting the most important points in "fact boxes" is one extremely effective way to engage readers while focusing on key takeaway points. 
This approach was developed to increase effective communication between stakeholders such as healthcare providers and patients (see for example: \url{https://www.hardingcenter.de}, archived in ). 
Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented on 14 March 2020 in the Washington Post\footnote{\url{https://www.washingtonpost.com/graphics/2020/world/corona-simulator/}} with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve".  Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge. 
While human thinking tends to simplify patterns, and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. 
Thus, in studying COVID-19 and predicting its spread, it is important not only to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results. 
Associations observed in the data can often be caused by third party variables (confounders). 
In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. 
However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.

@@

\subsection{Data Visualization Design and Testing}
Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what forms are used to represent the data? 
How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualization (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.

@@

\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country as well as by the number of tests performed and the accuracy of the recording and reporting system. 
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
\end{enumerate}

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
	\caption{Choropleth map of the Covid-19 cases world-wide by country. Source: WHO \url{https://covid19.who.int}}
	\label{fig:choro1}
\end{figure*}

@@

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
	\caption{Proportional symbol map showing the cumulated reported Covid-19 related deaths adjusted for population world-wide by country. Source: Washington Post \url{https://www.washingtonpost.com/graphics/2020/world/mapping-spread-new-coronavirus/}}
	\label{fig:propsymb}
\end{figure*}

@@

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. 
The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight Covid-19 is better? 
We will examine these topics more closely in Section~\ref{sec:rankings}.

\begin{figure}
\centering
\begin{subfigure}[c]{.45\textwidth}
\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state. The COVID Tracking Project \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
\end{subfigure}\hfill
\begin{subfigure}[c]{.45\textwidth}
\includegraphics[width=\textwidth]{china-dotmap}
\caption{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
\end{subfigure}
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
\caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways. Both methods are applied at the national scale rather than at global scale because of limitations with the visual form or data quality.}
\end{figure}

% \begin{figure}
% \centering
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
% \caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people. This somewhat mitigates the area bias issue, but is still subject to other perceptual biases. Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
% \end{subfigure}\hfill
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
% \caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways.}
% \end{figure}

@@

% Is it worth discussing e.g. https://www.nytimes.com/interactive/2021/us/covid-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem? When everything is always extremely high risk we just tend to tune things out. I don't want to distract from the overview, though... https://twitter.com/rypan/status/1282020934044401664 is a good example

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. 
Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig-jbm-tweet} and \Cref{fig:wapo-covid-time-series}. 
This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{wapo-covid-monitoring}
\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page. Source: \url{https://www.washingtonpost.com/coronavirus/}, 2022-01-17.}
\label{fig:wapo-covid-time-series}
\end{figure}

The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-covid-time-series}. This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life. Often, as in \Cref{fig:wapo-covid-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe. Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{wapo-covid-context}
\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots. Source: Washington Post Coronavirus Tracking, \url{https://www.washingtonpost.com/graphics/2020/national/coronavirus-us-cases-deaths/}.}
\label{fig:wapo-context}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ft_vaccine_bubble}
\caption{Animated bubble chart showing vaccination rate relative to per capita GDP for each country, from the Financial Times (Source: \url{https://ig.ft.com/coronavirus-vaccine-tracker})}\label{fig:bubble-ft}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss. Source: Romain Vuillemot, \url{https://observablehq.com/@romsson/sonification-of-covid-19-data}.}
\label{fig:freq-domain}
\end{figure}

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{Visual narratives of the COVID-19 pandemic}
\Plaintitle{Visual narratives of the COVID-19 pandemic}
\Shorttitle{Visual narratives of COVID-19}

%% - \Abstract{} almost as usual
\Abstract{
  COVID-19 has sparked a worldwide interest in understanding the dynamic evolution of a pandemic and tracking the effectiveness of preventive measures and rules. For this reason, numerous media and research groups have produced comprehensive data visualisations to illustrate the relevant trends and figures. In this paper, we will look at a selection of COVID 19 data visualisations to evaluate and discuss the currently established visualisation tools in terms of their ability to provide a communication channel both within the data science team and between data analysts, domain experts and a general interested audience. Although there is no set catalogue of evaluation criteria for data visualisations, we will try to give an overview of the different core aspects of visualisation evaluation and their competing principles.
  }

@@

Over the past two years, several waves of COVID-19 infections caused by different mutations of the SARS-CoV-2 virus have swept across the globe, claiming lives, stressing healthcare systems, and changing many different facets of our day-to-day experience. According to the WHO, almost 600 million confirmed cases and over 6.45 million deaths have been reported as of late August 2022 \footnote{\url{https://www.who.int/docs/default-source/coronaviruse/who_mou_30_august-2022.pdf}}. The pandemic has generated enormous interest in epidemiological data, its analysis and visualisation. From the beginning of the pandemic, data on the number of infections and COVID-related deaths has been published daily and made available to the public. Media, politicians and individuals use this data to build their narratives about the pandemic, discuss its evolution, justify the measures taken and discuss different prevention strategies against the spread of the virus. %So there are several goals to be achieved by visualising COVID-19 data.
These goals and their priorities were adapted as dynamically as the virus mutated and the pandemic changed pace, but as with any other data visualisation the general principles remain the same: ensuring clear understanding by organising graphics in such a way that the story of the data is told most effectively.

@@

In recent decades, many media organisations have established data teams that received unprecedented attention and wide-ranging opportunities during the pandemic to showcase their skills and abilities, not only in visualising data, but also in explaining their data collection and data analysis strategies and methods. COVID-19 allowed these data journalists to prove their utility in newsrooms, presenting data in digestible form to the public in news articles but also Data journalism will certainly be one of the beneficiaries of the COVID-19 pandemic and it has become an important part of news publishing, with COVID-19 delivering many excellent applications, often presented in an interactive visual format on the web, such as dashboards\citep{kochWelcomeRevolutionCOVID192021}.

@@

At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental statistical and visualisation design principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. 
These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others \citep{EQUATORNetworkEnhancing}. 
Efforts to improve the communication of quantitative and statistical information date from the beginning mass media and the ability to create charts and tables from data systematically, and while there continue to be efforts made in this area, historical papers outline many of the same issues that we still struggle with today \citep{ASA-standards,haemerPresentationProblemsArea1949,kovermanPLANNINGVISUALPRESENTATION1961,fienbergGraphicalMethodsStatistics1979,Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.
In other words, while our **technology** for recording, visually summarizing, and communicating information distilled from data has vastly improved over the past 100 years, we still face many of the same problems when **designing** charts which communicate data accurately and effectively.

@@

\subsection{Effective Data Communication}
A best practice for effective communication (graphical or otherwise) is to present information in an understandable way \citep{gigerenzerHelpingDoctorsPatients2007,gigerenzerWhatAreNatural2011}, for example by saying "one in ten" instead of 10\%, or using absolute rather than relative numbers. 
Presenting information in a visually appealing graphic form and highlighting the most important points in "fact boxes" is one extremely effective way to engage readers while focusing on key takeaway points. 
This approach was developed to increase effective communication between stakeholders such as healthcare providers and patients \citep{FaktenboxenZurMRNASchutzimpfung}. 
Such "fact boxes" combined with the dynamic layer of the internet allowed for illustrative simulations of the spread of the epidemic, for example as presented on 14 March 2020 in the Washington Post with the title "Why outbreaks like coronavirus spread exponentially, and how to flatten the curve" \citep{stevensTheseSimulationsShow2020}.  Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge. 
While human thinking tends to simplify patterns, and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. 
Thus, in studying COVID-19 and predicting its spread, it is important not only to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results. 
Associations observed in the data can often be caused by confounding variables. 
In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. 
However, statisticians who point out these limitations are at risk of having their statements pulled out to support one side in a polarised debate \citep{McConway2021}.

@@

\subsection{Data Visualization Design and Testing}
Having briefly discussed the media landscape surrounding the COVID pandemic and goals for communicating about data effectively, we should also briefly discuss the choices that go into creating visual representations of data intended to convey quantitative information.
While the overall goal of a data visualization is to tell the story of the data clearly and effectively, how this story is conveyed involves a series of design choices: which variables are shown, what limits are imposed, what visual forms are used to represent the data? 
How will the chart (and any surrounding story, in journalism) be presented to the viewer? 
As the designer makes these choices, they have in mind an audience, and the audience's assumed characteristics affect the design choices as well: language, education level, likely medium used to access the visualization (computer, physical newspaper, tablet, cell phone) - all of these considerations also impact the design of the chart and what ideas the designer considers during the creative process.

@@

\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Viewers tend to integrate similarly coloured areas of the map unconsciously, and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of COVID-19 cases are strongly influenced by the population size of the country as well as by the number of tests performed and the accuracy of the recording and reporting system. 
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of COVID cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Colour-scheme selection: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997, color2021}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.
\end{enumerate}

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/who_totalcases_choro.png}
	\caption{Choropleth map of the COVID-19 cases world-wide by country \citep{WHOCoronavirusCOVID19}.}
	\label{fig:choro1}
\end{figure*}

@@

\begin{figure*}
	\includegraphics[width = 0.98\textwidth]{Figures_Web/wp_totaldeaths_propsymb.png}
	\caption{Proportional symbol map showing the cumulated reported COVID-19 related deaths adjusted for population world-wide by country \citep{alcantaraMappingWorldwideSpread}.}
	\label{fig:propsymb}
\end{figure*}

@@

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe: none of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. 
The use of maps to assess the global pandemic more often focuses on the comparative aspects: how are we doing as opposed to our neighbors? Which strategy to fight COVID-19 is better? 
We will examine these topics more closely in \Cref{sec:rankings}.

\begin{figure}
\centering
\begin{subfigure}[c]{\textwidth}
\centering
\includegraphics[width=.65\textwidth]{COVID-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of COVID-19 cases in the United States by state \citep{StateCases1M2021}.}\label{fig:hex-cartogram}
\end{subfigure}

\begin{subfigure}[c]{\textwidth}
\centering
\includegraphics[width=\textwidth]{china-dotmap}
\caption{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
\end{subfigure}
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-COVID-scaled-choro}
% \caption{Choropleth Map of COVID cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/COVID-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
\caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways. Both methods are applied at the national scale rather than at global scale because of limitations with the visual form or data quality.}
\end{figure}
% 
% \begin{minipage}{\linewidth}
% \begin{minipage}{.7\linewidth}
% \includegraphics[width=\textwidth]{COVID-tracking-hex-cartogram}
% \end{minipage}
% \begin{minipage}{.29\linewidth}
% \captionof{figure}{Hexagonal Cartogram of COVID-19 cases in the United States by state. The COVID Tracking Project \url{https://COVIDtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
% \end{minipage}
% \end{minipage}
% 
% \begin{minipage}{\linewidth}
% \begin{minipage}{.7\linewidth}
% \includegraphics[width=\textwidth]{china-dotmap}
% \end{minipage}
% \begin{minipage}{.29\linewidth}
% \captionof{figure}{A dot-density map of COVID cases in China \citep{fieldMappingCoronavirusResponsibly}.}\label{fig:china-dot-map}
% \end{minipage}
% \end{minipage}

% \begin{figure}
% \centering
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{COVID-tracking-hex-cartogram}
% \caption{Hexagonal Cartogram of COVID-19 cases in the United States by state, per million people. This somewhat mitigates the area bias issue, but is still subject to other perceptual biases. Source: The COVID Tracking Project at The Atlantic \url{https://COVIDtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
% \end{subfigure}\hfill
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-COVID-scaled-choro}
% \caption{Choropleth Map of COVID cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/COVID-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
% \caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways.}
% \end{figure}

@@

% Is it worth discussing e.g. https://www.nytimes.com/interactive/2021/us/COVID-risk-map.html that does not make this distinction and runs into the "boy who cried wolf" problem? When everything is always extremely high risk we just tend to tune things out. I don't want to distract from the overview, though... https://twitter.com/rypan/status/1282020934044401664 is a good example

@@

Visualisations are typically created for a specific purpose, and during a pandemic it is particularly critical to show the evolution of the situation over time, as there are new developments on a daily basis. 
Typically, this means that cases, hospitalizations, and/or deaths are plotted on the y-axis, with date shown on the x-axis, as in \Cref{fig:jbm-tweet} and \Cref{fig:wapo-COVID-time-series}. 
This not only allows individuals to compare the present to past case counts, but also allows for forecasting and prediction of the future based on the current state. 
Time series plots are the most natural choice for these goals, and over the course of the pandemic, many different attempts have been made to show time series information using different graphical forms.

\begin{figure}
\centering
\includegraphics[width=.99\linewidth]{wapo-COVID-monitoring}
\caption{COVID monitoring summary charts on the Washington Post's Coronavirus page \citep{Coronavirus2022}.}
\label{fig:wapo-COVID-time-series}
\end{figure}

The most obvious time-series plot is also the most common: plotting cases against date, often with a 7 or 14-day smooth to handle periodicity in case reporting due to the work week, as shown in \Cref{fig:wapo-COVID-time-series}. This allows for individuals to assess the current situation and easily compare the current number of cases to past times for which the individual has a direct reference for what to expect in impact to daily life. Often, as in \Cref{fig:wapo-COVID-time-series}, these time series show multiple variables: cases, deaths, and hospitalizations (and sometimes vaccinations) to provide a more comprehensive view of the situation; this is particularly useful as there has been some decoupling of case numbers and hospitalizations/deaths with the emergence of effective vaccinations and strains which are more contagious but potentially also less severe. Many outlets also provide additional charts designed to provide some immediate context to readers, as shown in \Cref{fig:wapo-context}.

\begin{figure}
\centering
\includegraphics[width=.5\linewidth]{wapo-COVID-context}
\caption{Some outlets provide contextual information to assist readers with interpreting time-series plots \citep{CoronavirusCasesTracking2022}.}
\label{fig:wapo-context}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ft_vaccine_bubble}
\caption{Animated bubble chart showing vaccination rate relative to per capita GDP for each country \citep{visualCovid19VaccineTracker}.}\label{fig:bubble-ft}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.75\linewidth]{frequency-domain}
\caption{Frequency-domain case count graphics. Here, each 1000 deaths are represented by a single line segment whose slope represents the time taken to reach the next 1000 deaths. This provides viewers with a sense of the pace of the epidemic, rather than the raw case counts in more standard time-series representations. This representation can also be shown in the auditory domain, providing access to those who are typically excluded from visualizations due to vision loss \citep{vuillemotSonificationCOVID19Data2020}.}
\label{fig:freq-domain}
\end{figure}
